%\documentclass[12pt]{scrreprt}
\documentclass[12pt]{report}

% language may be romanian or english (default is english)
% type may be bachelor or master (default is bachelor)
\usepackage[language=english, type=bachelor]{style}

%controlling the appearance of your headers and footers
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\chead{}
\renewcommand{\headrulewidth}{0.2pt}
\renewcommand{\footrulewidth}{0.2pt}

\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray},
  stringstyle=\color{red!70!black},
  breaklines=true,
  frame=single,
  captionpos=b,
  tabsize=2
}

\begin{document}

\specialization{Computer Science}
\title{Design and Evaluation of a Trustworthy AI Shopping Assistant
for E-Commerce}
\author{Mihai}
\supervisor{[Title and name of supervisor]}

\maketitle

\newpage
\thispagestyle{empty}
\mbox{}
\newpage
\pagenumbering{roman}

\cleardoublepage
\textbf{\Large ABSTRACT}
\vspace{0.5cm}
\hrule
\vspace{0.5cm}

Online shopping systems increasingly integrate large language models
to support product discovery and decision-making. However, many AI
shopping assistants are intrusive, poorly timed, or perceived as
upselling tools rather than genuine advisors. This thesis presents
the design, implementation, and evaluation of a reactive-first AI
shopping advisor for e-commerce. The proposed system combines a
classical recommendation layer based on co-occurrence, content
similarity, and personalized ranking with a behavior-aware readiness
model and an LLM-based explanation layer. Instead of proactively
interrupting the user, the system uses lightweight browsing signals
to infer when help may be useful and only generates AI advice when
the user explicitly engages.

The system is implemented as a full-stack prototype with a simulated
e-commerce store, a seeded product and order dataset, a
recommendation backend, and an interactive advisor interface.
Evaluation is performed at three levels: recommendation quality,
system performance, and user perception. Offline recommender metrics
are used to assess the baseline recommendation layer, while latency
and token-based cost measurements capture the efficiency of the
AI-assisted pipeline. A structured user evaluation compares the
baseline shopping experience with the reactive-first AI-assisted
version in terms of perceived usefulness, trust, intrusiveness, and
purchase confidence.

The thesis aims to show that hybrid architectures can deliver useful
and explainable shopping assistance while avoiding the drawbacks of
overly proactive AI interaction. The results contribute practical
guidance for building trustworthy, non-intrusive AI assistants in
consumer-facing systems.

\tableofcontents

\newpage
\pagenumbering{arabic}

\input{chapters/chapter1_introduction}
\input{chapters/chapter2_background}
\input{chapters/chapter3_problem}
\input{chapters/chapter4_design}
\input{chapters/chapter5_implementation}
\input{chapters/chapter6_evaluation}
\input{chapters/chapter7_conclusions}

\bibliography{references}

\end{document}
